---
title: "HW 06: Logistic regression"
subtitle: "Binomial responses and overdispersion"
format: 
  pdf:
    fig-height: 4
    fig-width: 6
---

```{r}
#| label: load-packages

library(tidyverse)
library(broom)
library(knitr)

# add other packages as needed
```

## Data: Supporting railroads in the 1870s {.midi}

The data set [`RR_Data_Hale.csv`](data/RR_Data_Hale.csv) contains information on support for referendums related to railroad subsidies for 11 communities in Hale County, Alabama in the 1870s. The data were originally collected from the US Census by historian Michael Fitzgerald and analyzed as part of a thesis project by a student at St. Olaf College. The variables in the data are

-   `pctBlack`: percentage of Black residents in the county
-   `distance`: distance the proposed railroad is from the community (in miles)
-   `YesVotes`: number of "yes" votes in favor of the proposed railroad line
-   `NumVotes`: number of votes cast in the election

```{r}
#| label: load-data

rr <- read_csv("data/RR_Data_Hale.csv")

rr <- rr |>
  mutate(pctYes = YesVotes/NumVotes, 
         emp_logit = log(pctYes / (1 - pctYes)), 
         inFavor = if_else(pctYes > 0.5, "Yes", "No"))
```

# Part 1

```{r}
#| label: fit-model

rr_model <- glm(cbind(YesVotes, NumVotes - YesVotes) ~ distance + pctBlack, 
                data = rr, family = binomial)

tidy(rr_model, conf.int = TRUE) |>
  kable(digits = 3)
```

**Alternate model syntax**

```{r}
#| label: fit-model-alt-syntax

rr_model_alt <- glm(pctYes ~ distance + pctBlack, data = rr, 
                    family = binomial, weight = NumVotes)

tidy(rr_model_alt, conf.int = TRUE) |>
  kable(digits = 3)
```

::: callout-note
## Exercise 1

Interpret the coefficient of distance in the context of the data.
:::



For each mile increase in the distance the community is from the proposed railroad, the odds that the community voted Yes for the railroad changes by a multiplicative factor of `{r} exp(-0.292)` i.e., it declines by 25.3% after adjusting for racial composition.

::: callout-note
## Exercise 2

Use a likelihood ratio test or drop-in-deviance test to determine if the interaction between `distance` and `pctBlack` should be added to the model.
:::

drop in deviance
$$
H_0: distance, percentblack
$$


$$
H_1: distance, percentblack, dist\times percBlack
$$
```{r}
rr_model_int <- glm(pctYes ~ distance + pctBlack + distance*pctBlack, data = rr, 
                    family = binomial, weight = NumVotes)

tidy(rr_model_int, conf.int = TRUE) |>
  kable(digits = 3)
```




```{r}

anova(rr_model_alt, rr_model_int,test = "Chisq")

```

We perform a drop in deviance, and our p-value is very small so we reject the null in favor of the model with the interaction term. We find that there is significant evidence supporting the full model. 


::: callout-note
## Exercise 3

Use the model selected in the previous exercise. Interpret the effect of the demographics for a community that is...

-   Right on the proposed railroad (distance = 0)

-   15 miles away from the proposed railroad (distance = 15)
:::

For a community that is right on the railroad, when percent black increases by 1, the odds that the community voted Yes for the railroad changes by a multiplicative factor of `{r} exp(-0.065)` i.e. it decreases by about 6.3%.	


For a community 15 miles away from the proposed railroad, increase in odds in 1.5%.


::: callout-note
## Exercise 4

Conduct the appropriate test to assess if the model selected in Exercise 2 is good fit for the data.
:::

```{r}
1-pchisq(274.23, 7)
```

This is statistically significant evidence of lack of fit.
# Part 2

::: callout-note
## Exercise 5

Fit the quasibinomial model. How did the coefficients change from the original model? How did the standard errors change?
:::

```{r}

rr_model_qb <- glm(pctYes ~ distance + pctBlack + distance*pctBlack, data = rr, 
                    family = quasibinomial, weight = NumVotes)

tidy(rr_model_qb, conf.int = TRUE) |>
  kable(digits = 3)

```


The coefficients did not change but the standard errors were inflated.

::: callout-note
## Exercise 6

Based on the results from Exercise 5, what might be your next step in the analysis? If possible, conduct that step below.
:::

We noticed that the p-values are not below 0.05 in the quasibinomial fit, so I think we should start with a simple model again that just has distance. Then we can add percentBlack and compare. Then add the interaction term and compare.



```{r}

rr_model_qb_simple <- glm(pctYes ~ distance, data = rr, 
                    family = quasibinomial, weight = NumVotes)

tidy(rr_model_qb_simple, conf.int = TRUE) |>
  kable(digits = 3)

```


```{r}

rr_model_qb_two <- glm(pctYes ~ distance + pctBlack, data = rr, 
                    family = quasibinomial, weight = NumVotes)

tidy(rr_model_qb_two, conf.int = TRUE) |>
  kable(digits = 3)
```


```{r}
anova(rr_model_qb_simple, rr_model_qb_two, test = "Chisq")
```

